Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-07-26 submission/manuscript.Rmd
discussion update 2

A substantial portion of genes upregulated in _Kdm5c_-KO brain are testis genes that have no known function within the brain. Through the use of publically available RNA-seq datasets, we demonstrated these testis-enriched DEGs are typically unique to germ cells. Misexpression of germline genes in the brain suggests _Kdm5c_-KOs fail to demarcate between somatic and germline lineages, a key feature of multicellularity and sexual reproduction. Previous studies have demonstrated germline genes are decommissioned in somatic cells by multiple chromatin-modifying enzymes, but have primarily focused on the repression of key marker genes. To globally characterize KDM5C's role in germline gene repression, we curated a list of mouse germline-enriched genes using publically available germ cell-depleted RNA-seq datasets. This resource enabled us to identify 1) the extent of germline gene dysregulation in _Kdm5c_-KO cells, 2) the types of germline genes misexpressed at different developmental time points, and 3) which types of germline genes are directly or indirectly regulated by KDM5C.

@@

\clearpage
# Notes
## Dazl
We were particularly interested in the aberrant transcription of _Dazl_, since it is essential for germ cell development and promotes the translation of germline mRNAs[@liDAZLMasterTranslational2019]. A significant portion of germline transcripts misexpressed in _Kdm5c_-KO EpiLCs are known binding targets of DAZL, including _Stra8_[@mikedisDAZLMediatesBroad2020] (p = 1.698e-07, Fisher’s Exact Test). This suggests expression of DAZL protein could promote the translation of other aberrant germline transcripts, influencing their ability to impact _Kdm5c_-KO cellular function. We thus tested DAZL protein expression in _Kdm5c_-KO EpiLCs through immunocytochemistry (Figure 3I). We observed about 25% of _Kdm5c_-KO EpiLCs expressed DAZL protein and it was localized to the cytoplasm (p = 0.0015, Welch's t-test), consistent with the pattern of DAZL expression in spermatogonia[@mikedisDAZLMediatesBroad2020]. Altogether these results suggest tissue-specific genes are misexpressed during _Kdm5c_-KO embryogenesis, including key drivers of germline identity that can be translated into protein.

>>>>

We found a substantial portion of genes upregulated in _Kdm5c_-KO brain are germline genes with no known fucntions within the brain. Distinguishing the germline (i.e. cells that pass on their genetic material to the next generation) from the soma (i.e. all other cells of the body) is a key feature of multicellularity and sexual reproduction[XXX]. Previous work identifying chromatin regualtors that silence germ cell-specific transcription has predominatly focused on their repression of key marker genes, such as _Dazl_ and _Ddx4_[XXX]. To globally characterize KDM5C's role in germline gene repression, we curated a list of mouse germline-enriched genes using publically available germ cell-depleted RNA-seq datasets from Kit^W/Wv^ mice[]. This resource enabled us to identify 1) the extent of germline gene dysregulation in _Kdm5c_-KO cells, 2) the types of germline genes misexpressed at different developmental time points, and 3) the categories of germline genes directly and indirectly regulated by KDM5C.

@@

\clearpage
# Notes
## Things to do
- Move dazl to new figure if other staining works
- Use Ddx4 staining (with dazl?). If not should add RNA to diagram because its a key germline gene.
- Add cell culture concentrations to methods

## Dazl
We were particularly interested in the aberrant transcription of _Dazl_, since it is essential for germ cell development and promotes the translation of germline mRNAs[@liDAZLMasterTranslational2019]. A significant portion of germline transcripts misexpressed in _Kdm5c_-KO EpiLCs are known binding targets of DAZL, including _Stra8_[@mikedisDAZLMediatesBroad2020] (p = 1.698e-07, Fisher’s Exact Test). This suggests expression of DAZL protein could promote the translation of other aberrant germline transcripts, influencing their ability to impact _Kdm5c_-KO cellular function. We thus tested DAZL protein expression in _Kdm5c_-KO EpiLCs through immunocytochemistry (Figure 3I). We observed about 25% of _Kdm5c_-KO EpiLCs expressed DAZL protein and it was localized to the cytoplasm (p = 0.0015, Welch's t-test), consistent with the pattern of DAZL expression in spermatogonia[@mikedisDAZLMediatesBroad2020]. Altogether these results suggest tissue-specific genes are misexpressed during _Kdm5c_-KO embryogenesis, including key drivers of germline identity that can be translated into protein.
